\clearpage
\phantomsection
\addcontentsline{toc}{chapter}{ACKNOWLEDGEMENT}
\thispagestyle{footeronly}

\begin{center}
{\Large \textbf{ACKNOWLEDGEMENT}}
\end{center}

\vspace{0.5cm}

\noindent
At the outset, we sincerely thank the Almighty for granting us the strength, wisdom, and perseverance to successfully complete this project.

\vspace{0.3cm}

\noindent
We would like to convey our heartfelt gratitude to our respected project guide, \textbf{Ms. Swaiba Nasmi}, whose expert guidance, thoughtful suggestions, and continuous encouragement greatly influenced the direction and quality of our work. Her readiness to clarify our doubts and provide valuable feedback at every stage played a crucial role in the completion of this project.

\vspace{0.3cm}

\noindent
We are also deeply thankful to \textbf{Dr. Pramod Pavithran}, Head of Division, for his constant motivation, academic support, and for providing us with the necessary facilities to carry out this work successfully.

\vspace{0.3cm}

\noindent
We extend our sincere appreciation to \textbf{Dr. Sheena S}, Staff Coordinator, for her constant support, guidance, and encouragement throughout the project duration. Her valuable insights and suggestions helped us improve and refine our implementation.

\vspace{0.3cm}

\noindent
Our sincere appreciation extends to our classmates and friends who supported us with their ideas, cooperation, and encouragement throughout the development process. Their contributions helped us improve and refine our implementation.

\vspace{0.3cm}

\noindent
Finally, we express our deepest gratitude to our families for their unconditional support, patience, and belief in our abilities. Their encouragement has always been the foundation of our achievements.

\vspace{1.5cm}
\noindent
Nema Fathima (20223071)\\
Nidha Rahma (20223078)\\
Rithuparna J B (20223084)\\
Rose Maria (20223085)\\
Sana Noushad (20223092)

\vfill
\clearpage