\chapter{Introduction}
\label{ch:introduction}

This chapter introduces the Automated Diagram Digitization project, 
presenting the motivation behind the work, the key problems it 
addresses, its primary objectives, and the scope of the system 
developed.

In today's digital era, the need to convert hand-drawn or printed 
diagrams into editable digital formats has become increasingly 
important across various domains including education, software 
development, and business process modeling. Flowcharts, being 
fundamental tools for representing algorithms, processes, and 
workflows, are frequently created on paper or whiteboards during 
brainstorming sessions, classroom teachings, and initial design 
phases. However, these hand-drawn diagrams often remain in physical 
form, limiting their usability, shareability, and 
editability.

Traditional methods of digitizing flowcharts involve manual redrawing 
using diagramming tools like Microsoft Visio, Lucidchart, or draw.io 
-- a time-consuming and error-prone process. This creates a 
significant gap between the creative, free-form sketching process and 
the structured, digital representation required for documentation and 
further processing. Recent advances in artificial intelligence, 
particularly in computer vision and large language models, have opened 
new possibilities for automated diagram recognition, enabling the 
development of intelligent systems that can understand and reconstruct 
diagrammatic information automatically.

%---------------------------------------------------------------
\section{Problem Statement}
%---------------------------------------------------------------

While manual diagramming tools have evolved significantly, the 
transition from physical brainstorming on whiteboards or paper to 
digital documentation remains a bottleneck in the software development 
lifecycle. Current workflows often require developers and designers to 
spend valuable time recreating existing work, leading to decreased 
productivity and potential inconsistencies between the original sketch 
and the digital version. This gap highlights the need for a more 
intelligent, automated bridge that can interpret visual structures 
without exhaustive manual input. Specifically, the existing diagram digitization ecosystem suffers from a series of interrelated challenges, starting with the significant manual redrawing bottleneck where users must recreate diagrams in tools like Visio or draw.io -- a process that often takes 15--30 minutes and introduces transcription errors. This problem is compounded by a lack of automated recognition for shapes and connections, as well as the failure of standard OCR tools to handle handwritten text or preserve line breaks within diagrammatic structures. Moreover, current systems struggle to identify directional relationships and offer limited input flexibility, often lacking support for AI-generated prompts or structured JSON data. These limitations are further exacerbated by the absence of intelligent editing features like hierarchical auto-layout and restricted export formats, which collectively hinder the seamless transition from visual sketches to maintainable digital assets.

%---------------------------------------------------------------
\section{Objectives}
%---------------------------------------------------------------

The project aims to overcome the limitations of traditional digitization methods by providing a comprehensive, AI-driven platform that simplifies the entire flowchart lifecycle—from initial recognition to final export. By integrating state-of-the-art vision models with a highly interactive editing environment, the system seeks to empower users with tools that were previously only available through complex, expensive enterprise software. The Automated Diagram Digitization project focuses on developing an AI-powered recognition system capable of identifying flowchart shapes, text, and connections from images using Google's Gemini AI with custom prompt engineering for accurate shape detection. It provides multiple input methods, including image upload, text-based algorithm-to-flowchart conversion, and JSON import, ensuring maximum flexibility and seamless workflow integration. The system includes an intuitive interactive diagram editor with drag-and-drop functionality, undo/redo support, auto-layout for proper alignment, and comprehensive styling options such as node and edge color customization. Additionally, it enables advanced text formatting features like bold, italic, and font size adjustments to enhance readability and presentation quality. Edge labeling functionality is implemented through double-click editing, allowing users to add and modify connection labels, particularly for decision branches such as ``Yes/No'' with proper styling. The platform also supports multiple export formats including PNG for sharing, SVG for high-quality vector graphics, and JSON for data portability and integration with other tools. Theme support is provided through a dark/light mode toggle with persistent user preferences stored in \texttt{localStorage} to improve user experience and reduce eye strain. Furthermore, the system ensures persistent storage by enabling saving and retrieval of flowcharts using a PostgreSQL database with UUID-based identification and timestamp tracking for audit purposes.

%---------------------------------------------------------------
\section{Scope of the Project}
%---------------------------------------------------------------

The Automated Diagram Digitization project is designed with a well-defined scope that guides development priorities and sets clear expectations for its capabilities. The scope is categorized into functional and technical aspects, each addressing different dimensions of the system.

\subsection{Functional Scope}

The functional scope encompasses all end-user-visible features and system behaviors. The system transforms raw visual inputs into interactive digital representations while preserving the original logical structure of diagrams. It includes the ability to identify fundamental flowchart shapes such as process (rectangle), decision (diamond), start node (oval), and input/output (parallelogram) from uploaded images. It ensures accurate text extraction from shapes with proper preservation of line breaks using \texttt{\textbackslash n} characters for multi-line labels. The system performs automatic detection of connections between nodes, including directional arrows and appropriate handle assignments such as top, bottom, left, and right. It supports image input in formats like PNG and JPG, along with a drag-and-drop interface and file size validation. Additionally, it enables text-to-flowchart conversion using natural language processing to transform algorithm descriptions into structured diagrams with correct node types and connections. JSON import and export functionality allows users to load existing flowcharts for editing and export current states for portability. The interactive editor provides drag-and-drop capabilities from a shape palette, node resizing with adaptive text, undo/redo functionality with keyboard shortcuts, and auto-layout for clean alignment. Styling features include node and edge color customization, text formatting such as bold, italic, and font size adjustments, as well as edge labeling with styled backgrounds for improved readability.

\subsection{Technical Scope}

The technical scope defines the technologies and implementation strategies used to build the platform in a reliable and maintainable manner. The backend is implemented using FastAPI in Python with asynchronous endpoints supporting image analysis, text-to-flowchart generation, and database operations. The frontend is developed using React 19 with Vite for optimized performance, along with React Flow for diagram rendering and custom node components tailored for each shape type. The system integrates with the Google Gemini AI API using carefully designed system prompts to achieve accurate shape recognition and text-to-flowchart conversion. A PostgreSQL database is used with JSONB support to store the complete React Flow state, utilizing UUID-based primary keys and timestamp tracking. Export functionality is implemented using libraries such as html-to-image for PNG generation, React Flow’s SVG renderer for vector exports, and proper serialization for JSON output. The platform also follows a RESTful API design, providing endpoints for image analysis, text generation, and full flowchart CRUD operations.


\subsection{Limitations and Exclusions}

The project does not cover recognition of complex BPMN symbols, 
swimlanes, UML diagrams, or real-time video processing. It focuses 
specifically on the four basic flowchart shapes commonly used in 
algorithm representation and process documentation. Support for 
additional diagram types, collaborative editing, and mobile 
application interfaces are identified as directions for future 
development and are discussed in Chapter~8.

%---------------------------------------------------------------
\section{Significance of Work}
\label{sec:significance}
%---------------------------------------------------------------

The Automated Diagram Digitization project addresses a genuine and 
widespread problem faced by students, educators, and software 
professionals who regularly work with hand-drawn or printed 
flowcharts. Traditionally, converting such diagrams into editable 
digital formats required significant manual effort using tools like 
Microsoft Visio, Lucidchart, or draw.io -- a process that could take 
anywhere from 15 to 30 minutes per diagram.

By leveraging the multimodal capabilities of Google's Gemini AI 
combined with a purpose-built interactive editor, this project 
demonstrates that the digitization process can be reduced to a matter 
of seconds with a high degree of accuracy. The platform's support for 
multiple input methods -- image upload, natural language description, 
and JSON import -- makes it adaptable to a wide range of real-world 
workflows.

Beyond individual productivity, the system contributes to the broader 
goal of making diagram creation and documentation more accessible. It 
reduces the skill barrier associated with professional diagramming 
tools and opens up accurate, editable flowchart generation to users 
with minimal technical background. The open-source technology stack 
ensures that the platform can be freely adopted, extended, and built 
upon by future developers and researchers in the field of intelligent 
document processing.

%---------------------------------------------------------------
% Chapter Summary and Transition
%---------------------------------------------------------------

The remainder of this report is organized as follows. Chapter~2 
presents a critical review of related literature spanning diagram 
recognition, handwriting analysis, AI-powered flowchart generation, 
and interactive diagramming tools. Chapter~3 discusses the system 
study, including an analysis of existing systems, feasibility 
assessment, and requirements specification. Chapter~4 presents the 
detailed system design covering architecture, data flow, database 
schema, and API design. Chapter~5 describes the implementation of 
each module. Chapter~6 covers the technologies and tools used in 
development. Chapter~7 presents the testing strategy and results. 
Chapter~8 outlines potential future enhancements, and Chapter~9 
concludes the report with a summary of findings and reflections on 
the project.