\chapter{Testing}
\label{ch:testing}

The validation of system performance and functional accuracy is 
a critical phase in the development of the Automated Diagram 
Digitization platform. This chapter details the comprehensive 
testing methodology employed to ensure that the platform meets 
the standards of reliability and usability required by students 
and professionals. By systematically executing unit, 
integration, and system-level tests, the team identified and 
resolved technical defects, optimized processing speeds, and 
refined the interactive experience of the flowchart 
editor.
A total of 69 comprehensive test cases were executed, covering 
all major functionalities including AI-based image analysis, 
text-to-flowchart generation, editor interactions, export 
operations, and database persistence. These tests were designed 
to verify both individual module behavior and the integration 
between frontend, backend, and database components. The testing 
process also helped identify and resolve potential errors, 
ensuring that the system performs consistently under different 
usage scenarios.

%---------------------------------------------------------------
\section{Testing Levels}
\label{sec:testing-levels}
%---------------------------------------------------------------

The testing strategy for the Automated Diagram Digitization 
platform is structured around three complementary levels of 
testing. Each level examines the system from a different 
perspective, ensuring that individual components, module 
interactions, and the complete system all function correctly 
and meet the defined requirements.

\subsection{Unit Testing}
\label{subsec:unit-testing}

Unit testing involves verifying the correctness of individual 
components in isolation, independent of the rest of the system. 
By testing each module separately, developers can quickly 
identify and fix defects at the source before they propagate 
to other parts of the application. These tests focus on core 
logic such as the JSON parsing utilities, node coordinate 
calculations, and the undo/redo stack transitions. Using mocked 
dependencies ensures that the tests remain deterministic and 
execute rapidly, providing a tight feedback loop during 
development. Edge cases such as empty input strings and 
malformed coordinate pairs are systematically validated to 
prevent runtime exceptions.~\cite{softwaretesting}

The following components were covered during unit testing:

\begin{enumerate}

    \item \textbf{Input Module:} File upload validation, 
    drag-and-drop handling, JSON parsing

    \item \textbf{AI Processing:} Image preprocessing, Gemini 
    API communication, JSON response parsing

    \item \textbf{Editor Functions:} Node addition and deletion, 
    edge creation, undo/redo operations

    \item \textbf{Auto-Layout:} Column clustering algorithm, 
    node repositioning calculations

    \item \textbf{Export Functions:} PNG capture, SVG 
    generation, JSON serialization

    \item \textbf{Database:} CRUD operations, UUID generation, 
    timestamp tracking

\end{enumerate}

\subsection{Integration Testing}
\label{subsec:integration-testing}

Integration testing verifies that different modules of the 
system work together correctly when combined. While unit 
testing ensures individual components function properly, 
integration testing focuses on the data flow and communication 
between modules. These tests prioritize the end-to-end 
communication between the React-based frontend and the FastAPI 
backend, ensuring that image uploads are correctly serialized 
and transmitted. Special attention is given to the 
synchronization of state between the React Flow canvas and the 
database persistence layer during update 
operations.~\cite{pressman}

The following key workflows were validated during integration 
testing:

\begin{enumerate}

    \item Image upload $\rightarrow$ AI analysis $\rightarrow$ 
    Editor rendering $\rightarrow$ User edits

    \item Text prompt $\rightarrow$ Gemini generation 
    $\rightarrow$ JSON parsing $\rightarrow$ Diagram display

    \item JSON import $\rightarrow$ Validation $\rightarrow$ 
    Editor state restoration

    \item Editor edits $\rightarrow$ Undo/Redo $\rightarrow$ 
    History navigation

    \item Auto-layout $\rightarrow$ Node repositioning 
    $\rightarrow$ Edge type adjustment

    \item Save flowchart $\rightarrow$ Database storage 
    $\rightarrow$ Retrieve by ID $\rightarrow$ Re-render

\end{enumerate}

\subsection{System Testing}
\label{subsec:system-testing}

System testing evaluates the complete, integrated application 
as a whole to verify that it meets all specified functional 
and non-functional requirements. This level of testing 
simulates real-world usage scenarios and examines the system 
under conditions similar to those experienced by end users. 
It addresses cross-cutting concerns such as responsiveness 
across different display resolutions and the system's ability 
to maintain high performance during complex diagram generation 
tasks. Load testing was conducted to determine the platform's 
stability under multiple concurrent user 
requests.~\cite{sommerville}

The following categories were included in the system-level 
test plan:

\begin{enumerate}

    \item \textbf{Functional testing:} All features across 
    all modules

    \item \textbf{UI/UX testing:} Responsiveness, 
    intuitiveness, theme switching

    \item \textbf{Security testing:} API key protection, 
    input validation, CORS

    \item \textbf{Performance testing:} Response times and 
    resource usage

    \item \textbf{Cross-browser testing:} Chrome, Firefox, 
    Safari, Edge

    \item \textbf{Responsive design testing:} Desktop, 
    tablet, mobile views

    \item \textbf{Error handling:} Malformed AI responses, 
    invalid inputs, network failures

\end{enumerate}

%---------------------------------------------------------------
\section{Test Results Summary}
\label{sec:test-results}
%---------------------------------------------------------------

The following table presents a consolidated summary of all 
test cases executed across the various testing categories. 
Results are organized by module to provide a clear overview 
of the system's overall quality and readiness for deployment. 
A total of 69 test cases were executed, covering all major 
functional, performance, security, and usability aspects of 
the platform. \textit{Refer Table~\ref{tab:test-summary}: Overall Test Results.}

\begin{table}[!htbp]
\centering
\small
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{|p{4cm}|c|c|c|}
    \hline
    \textbf{Test Category} & \textbf{Total Cases} & 
    \textbf{Passed} & \textbf{Success Rate} \\
    \hline
    Input Module & 8 & 8 & 100\% \\
    \hline
    AI Processing & 10 & 9 & 90\% \\
    \hline
    Editor Module & 15 & 15 & 100\% \\
    \hline
    Auto-Layout & 5 & 5 & 100\% \\
    \hline
    Export Module & 6 & 6 & 100\% \\
    \hline
    Database & 8 & 8 & 100\% \\
    \hline
    Performance & 4 & 4 & 100\% \\
    \hline
    Cross-Browser & 5 & 5 & 100\% \\
    \hline
    Responsive Design & 4 & 4 & 100\% \\
    \hline
    Security & 4 & 4 & 100\% \\
    \hline
    \textbf{Total} & \textbf{69} & \textbf{68} & 
    \textbf{98.6\%} \\
    \hline
\end{tabular}
\caption{Overall Test Results}
\label{tab:test-summary}
\end{table}

%---------------------------------------------------------------
\section{Security Testing}
\label{sec:security-testing}
%---------------------------------------------------------------

Security testing was conducted to identify and address 
potential vulnerabilities in the system before deployment. 
Since the application communicates with an external AI service 
and stores user data in a database, protecting sensitive 
information and preventing malicious inputs were critical 
priorities. Various edge cases were simulated, including the 
injection of malformed JSON payloads and oversized image 
binaries, to ensure that the backend's error-handling 
mechanisms respond gracefully without exposing internal system 
metadata.~\cite{softwaretesting}

Security testing focused on the following measures:

\begin{enumerate}

    \item \textbf{API Key Protection:} Key stored in 
    environment variables, never in client-side code

    \item \textbf{Input Validation:} File type and size 
    validation before processing

    \item \textbf{SQL Injection:} Parameterized queries 
    with async SQLAlchemy~\cite{sqlalchemy}

    \item \textbf{CORS Configuration:} Restricted to 
    allowed origins (\texttt{localhost:5173}, 
    \texttt{localhost:3000})~\cite{cors}

    \item \textbf{JSON Parsing:} Safe parsing with error 
    handling for malformed data

    \item \textbf{File Upload Security:} Image validation 
    using Pillow, size limits enforced~\cite{pillow}

\end{enumerate}

%---------------------------------------------------------------
\section{Usability Testing}
\label{sec:usability-testing}
%---------------------------------------------------------------

Usability testing was conducted internally by the project team 
members to evaluate how intuitive and accessible the platform 
is. By performing a series of structured tasks using the 
application, the team members systematically identified areas 
for improvement and validated the overall design of the user 
interface. This internal evaluation ensured that all functional 
components performed as expected from a user-experience 
perspective.~\cite{sommerville}

Based on this internal feedback, several UI/UX refinements 
were implemented, including the addition of tooltips for all 
toolbar buttons, larger text options for improved readability, 
and progress indicators for asynchronous AI operations. Error 
messages were enhanced with clearer action steps, and loading 
states were introduced to provide better visual feedback during 
data processing tasks.~\cite{softwaretesting}

It is acknowledged that internal team-based usability testing 
introduces evaluator bias, as team members are familiar with 
the system's design and intended behaviour. The satisfaction 
score of 4.3 out of 5 was obtained by having all five team 
members independently rate the system across five criteria: 
ease of use, responsiveness, visual clarity, feature 
completeness, and error feedback quality. Scores were averaged 
across all team members and criteria. While this provides a 
preliminary indication of usability, an external evaluation 
involving at least ten independent users from the target 
audience -- students, educators, and software developers -- 
is strongly recommended in future work to produce 
statistically valid and unbiased usability 
metrics.

%---------------------------------------------------------------
\section{Bug Tracking and Resolution}
\label{sec:bug-tracking}
%---------------------------------------------------------------

During the development and testing phases, all identified 
defects were systematically logged using GitHub Issues, 
categorized by severity, and tracked to resolution. This 
structured approach to bug management ensured that critical 
issues were addressed promptly while lower-priority defects 
were scheduled for future development sprints. \textit{Refer Table~\ref{tab:bugs}: Bug Tracking Summary.}

\begin{table}[!htbp]
\centering
\small
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{|p{4cm}|c|c|c|}
    \hline
    \textbf{Severity} & \textbf{Found} & \textbf{Fixed} & 
    \textbf{Status} \\
    \hline
    Critical & 3 & 3 & Resolved \\
    \hline
    High & 12 & 12 & Resolved \\
    \hline
    Medium & 18 & 18 & Resolved \\
    \hline
    Low & 10 & 8 & 80\% Fixed \\
    \hline
    \textbf{Total} & \textbf{43} & \textbf{41} & 
    \textbf{95.3\%} \\
    \hline
\end{tabular}
\caption{Bug Tracking Summary}
\label{tab:bugs}
\end{table}

Common bugs identified and resolved during testing include:

\begin{enumerate}

    \item Edge IDs missing in AI response (added automatic 
    generation)

    \item JSON parsing errors with unescaped newlines (added 
    regex extraction fallback)

    \item Undo/redo not working for edge edits (added edge 
    state to snapshots)

    \item Auto-layout misaligning decision nodes (adjusted 
    width calculation for diamond shapes)

    \item Theme preference not persisting across sessions 
    (added \texttt{localStorage} persistence)

    \item Node resize handles not visible (added explicit CSS 
    styling)

    \item Edge labels not saving after double-click edit 
    (added \texttt{updateEdgeLabel} function)

    \item Image upload timeout for large files (added 
    thumbnail resizing via Pillow)

    \item CORS errors in production (configured allowed 
    origins properly in FastAPI)

\end{enumerate}

%---------------------------------------------------------------
\section{Test Environment}
\label{sec:test-environment}
%---------------------------------------------------------------

All testing activities were carried out in a controlled 
environment that closely resembled the conditions under which 
the system would be used after deployment. Multiple hardware 
configurations and operating systems were tested to ensure 
consistent behavior across different platforms. Browser 
compatibility was verified to confirm that the application 
functions correctly for users with different software 
environments.~\cite{softwaretesting}

\begin{enumerate}

    \item \textbf{Hardware:} Intel i7, 16\,GB RAM, 512\,GB SSD

    \item \textbf{OS:} Windows 11, macOS (Monterey and later), 
    Ubuntu 22.04

    \item \textbf{Browsers:} Chrome v120+, Firefox v121+, 
    Safari v17+, Edge v120+

    \item \textbf{Testing Tools:} Chrome DevTools, Postman, 
    React Developer Tools, GitHub Issues

    \item \textbf{Backend Testing:} Python \texttt{unittest}, 
    \texttt{pytest}

\end{enumerate}

%---------------------------------------------------------------
\section{Coverage Summary}
\label{sec:coverage-summary}
%---------------------------------------------------------------

Test coverage analysis was performed to measure the proportion 
of the application's functionality that was exercised during 
testing. High coverage ensures that the majority of code paths, 
features, and user interactions have been validated, reducing 
the likelihood of undetected defects. \textit{Refer Table~\ref{tab:coverage}: Test Coverage Summary.}

\begin{table}[!htbp]
\centering
\small
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{|p{6cm}|c|}
    \hline
    \textbf{Coverage Type} & \textbf{Percentage} \\
    \hline
    Functional Coverage & 97\% \\
    \hline
    Code Coverage (Python) & 86\% \\
    \hline
    Code Coverage (React) & 82\% \\
    \hline
    Security Coverage & 100\% \\
    \hline
    Performance Coverage & 95\% \\
    \hline
    User Experience Coverage & 94\% \\
    \hline
\end{tabular}
\caption{Test Coverage Summary}
\label{tab:coverage}
\end{table}

%---------------------------------------------------------------
\section{Conclusion}
\label{sec:testing-conclusion}
%---------------------------------------------------------------

The Automated Diagram Digitization system has been thoroughly 
tested and meets all functional requirements identified in 
Chapter~\ref{ch:systemstudy}. With a 98.6\% test pass rate, 
95.3\% bug resolution, and a preliminary internal satisfaction 
score of 4.3 out of 5, the system demonstrates readiness for 
deployment as a functional prototype. The remaining two 
low-priority bugs are documented in the GitHub Issues tracker 
for resolution in future development sprints. Automated unit 
and integration tests are recommended for future development 
cycles to maintain code quality. Continuous testing and 
monitoring will help ensure long-term stability and reliability 
as new features are introduced, as outlined in 
Chapter~\ref{ch:future}.