\chapter{Technologies Used}
\label{ch:technologies}

The selection of a modern and robust technology stack is a 
fundamental prerequisite for building a high-performance system 
like Automated Diagram Digitization. This chapter provides a 
detailed examination of the frameworks, libraries, and tools 
utilized in the development of the platform, outlining the 
technical rationale behind each choice. The architecture is 
designed to be modular, ensuring that the frontend interface, 
backend logic, and AI processing services remain loosely coupled 
for maximum scalability and maintainability.

The development of this platform leverages a JavaScript 
and Python ecosystem, combining the agility of React for the user 
interface with the high performance of FastAPI for backend 
services. By integrating Google's state-of-the-art Gemini AI 
models for multimodal analysis and PostgreSQL for reliable data 
persistence, the system achieves a balance between advanced 
intelligent capabilities and enterprise-grade stability. These 
technologies collectively enable the platform to handle complex 
diagram recognition tasks while providing a smooth, responsive 
experience for the end user.

%---------------------------------------------------------------
\section{Introduction}
\label{sec:tech-intro}
%---------------------------------------------------------------

Automated Diagram Digitization leverages modern web technologies 
to create an intelligent flowchart recognition and editing 
platform. This chapter summarizes the key technologies, 
frameworks, and tools employed in the project. \textit{Refer Table~\ref{tab:tech-stack}: Technology Stack Overview.}

\begin{table}[H]
\centering
\renewcommand{\arraystretch}{1.3}
\small
\begin{tabular}{|l|p{9cm}|}
    \hline
    \textbf{Component} & \textbf{Technologies} \\
    \hline
    \hline
    Frontend & React 19, Vite, React Flow, Custom Snap-Layout, 
               Vanilla CSS, Lucide React, React Icons \\
    \hline
    Backend & Python 3.9+, FastAPI, Uvicorn \\
    \hline
    Database & PostgreSQL 16, SQLAlchemy (async), JSONB \\
    \hline
    Containerization & Docker, Docker Compose \\
    \hline
    AI Integration & Google Gemini API, Pillow \\
    \hline
    Tools & VS Code, Git, GitHub, Swagger UI, npm \\
    \hline
    Dependencies & react-router-dom, python-dotenv, pydantic \\
    \hline
\end{tabular}
\caption{Technology Stack Overview}
\label{tab:tech-stack}
\end{table}

%---------------------------------------------------------------
\section{Frontend Technologies}
\label{sec:frontend-tech}
%---------------------------------------------------------------

The frontend of the Automated Diagram Digitization system is 
built using a modern JavaScript technology stack. 
The following subsections describe each frontend technology 
and its role within the application.

\subsection{React 19 with Vite}
\label{subsec:react-vite}

React 19 provides the component-based architecture for building 
reusable UI elements. Vite serves as the build tool, offering 
fast development server startup with hot module replacement and 
optimized production builds.

\subsection{React Flow}
\label{subsec:reactflow}

React Flow is the core diagramming library used for rendering 
interactive flowcharts. Key features utilized include custom 
node rendering, connection handles, node resizing, edge styling, 
and background grid with zoom and pan controls.

\subsection{Vanilla CSS, Lucide React, and React Icons}
\label{subsec:css-lucide}

Vanilla CSS provides a flexible and standards-based approach to styling the user interface with modern design patterns. Lucide React and React Icons supply a comprehensive library of modern SVG icons for the shape palette, toolbar buttons, and UI controls, ensuring a professional and intuitive visual language throughout the application.

\subsection{Custom Snap-Layout and React Router}
\label{subsec:layout-router}

\justifying
The Custom Snap-Layout algorithm serves as the primary 
organisational engine within the editor, providing the 
clustering models necessary for the auto-layout feature. 
React Router is utilized for client-side navigation, 
managing the seamless transition between the diagram input 
dashboard and the interactive editor interface.

%---------------------------------------------------------------
\section{Backend Technologies}
\label{sec:backend-tech}
%---------------------------------------------------------------

The backend services of the platform are implemented using a 
Python-based technology stack optimized for high-performance 
asynchronous processing, database interaction, and AI 
integration. The following subsections detail each 
component.

\subsection{FastAPI Framework}
\label{subsec:fastapi}

FastAPI serves as the backend framework, offering high 
performance comparable to Node.js with automatic OpenAPI 
documentation. The backend architecture leverages several core features of FastAPI to ensure high performance and developer productivity. The system utilizes asynchronous endpoint handling with the \texttt{async}/\texttt{await} pattern to manage concurrent requests efficiently. To maintain data integrity, the platform implements automatic request validation using Pydantic models, which enforce schema compliance for all incoming payloads. Furthermore, a built-in CORS middleware configuration ensures secure communication between the frontend and backend, while standardized JSON response serialization provides consistent error formatting and status reporting across all API endpoints.

\subsection{PostgreSQL with JSONB}
\label{subsec:postgresql}

PostgreSQL 15 provides reliable data persistence with JSONB 
support for flexible schema storage. The flowcharts table stores 
the complete React Flow state (nodes and edges) in a JSONB 
column, allowing schema-less storage of varying node 
structures.

\subsection{SQLAlchemy (Async)}
\label{subsec:sqlalchemy}

SQLAlchemy is a powerful and flexible ORM (Object-Relational 
Mapper) for Python that simplifies database interactions through 
high-level abstractions. The async version is used here to align 
with FastAPI's asynchronous processing model, ensuring that 
database queries do not block the event loop and allowing the 
backend to handle multiple concurrent requests 
efficiently. Security in the database layer is enhanced through the use of parameterized queries, which effectively prevent SQL injection attacks. Additionally, the schema employs UUID primary keys for secure, globally unique record identification and includes comprehensive timestamp tracking with full timezone support.

%---------------------------------------------------------------
\section{AI Integration}
\label{sec:ai-integration}
%---------------------------------------------------------------

The AI capabilities of the platform are powered by Google's 
Gemini multimodal model, supported by the Pillow image 
processing library for preprocessing input images before 
submission to the API. The subsections below outline how each 
component contributes to intelligent diagram 
analysis.

\subsection{Google Gemini API}
\label{subsec:gemini}

Gemini AI provides the multimodal intelligence for flowchart recognition and generation. The API is accessed via the official Python client library (\texttt{google-genai}) with custom system prompts engineered for shape detection and text extraction. Custom prompt engineering with the Gemini-Flash model achieves approximately 92\% accuracy in informal testing across a diverse sample set of 3 complex flowchart images (algorithmic, mathematical, and logical). Accuracy is evaluated based on node-type identification, text label precision, and edge connectivity correctness.~

\subsection{Pillow (PIL)}
\label{subsec:pillow}

Pillow handles image processing tasks including opening uploaded 
images, resizing to thumbnails (1024$\times$1024 pixels 
maximum), and format conversion before sending images to the 
Gemini API for analysis. This preprocessing step reduces API 
payload size and improves response latency.

%---------------------------------------------------------------
\section{Development Tools}
\label{sec:dev-tools}
%---------------------------------------------------------------

The development process was supported by a set of 
well-established tools for version control, API testing, and 
environment management. These tools ensured a structured, 
collaborative, and consistent development workflow throughout 
the project lifecycle.

\subsection{Version Control}
\label{subsec:version-control}

Git with GitHub enables systematic code management using feature 
branches and pull requests. The repository maintains separate 
directories for backend and frontend code, enabling parallel 
development and conflict-free 
collaboration.

\subsection{API Testing}
\label{subsec:api-testing}

The built-in Swagger UI provided by FastAPI is used extensively for 
testing and validating RESTful API endpoints during backend 
development, providing an interactive interface for exploring all system 
endpoints.

\subsection{Development Environment}
\label{subsec:dev-environment}

The Automated Diagram Digitization system was developed using 
a modern web development environment designed to support 
efficient frontend and backend development. The backend services 
were implemented using Python with the FastAPI framework, while 
the frontend interface was developed using React with Vite. Node.js and npm were used for managing frontend dependencies and running the development server. Python virtual environments were used to isolate backend dependencies and maintain a consistent 
development setup.

Version control was managed using Git, with the project 
repository hosted on GitHub, enabling collaborative development, 
version tracking, and seamless integration with cloud deployment 
platforms.Database schema design and data verification were facilitated using pgAdmin, 
and the built-in Swagger UI provided by FastAPI served as a 
critical tool for interactive API testing before frontend 
integration. \textit{Refer Table~\ref{tab:dev-tools}: Development Environment Tools}.

\begin{table}[H]
\centering
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{|l|p{7cm}|}
\hline
\textbf{Tool / Platform} & \textbf{Purpose} \\
\hline
Visual Studio Code & Code development and debugging \\
\hline
Node.js and npm & Frontend dependency management \\
\hline
Python Virtual Environment & Backend dependency isolation \\
\hline
Git and GitHub & Version control and collaboration \\
\hline
Swagger UI / Browser Testing & API testing and validation \\
\hline
\end{tabular}
\caption{Development Environment Tools}
\label{tab:dev-tools}
\end{table}

%---------------------------------------------------------------
\section{Project Structure}
\label{sec:project-structure}
%---------------------------------------------------------------

A well-organized project structure is essential for maintaining 
code clarity and enabling efficient collaboration among team 
members. The Automated Diagram Digitization project follows a 
clean separation between the frontend and backend directories, 
ensuring that developers can work on individual components 
independently.
\begin{enumerate}

    \item \textbf{backend/} -- FastAPI application
    \begin{enumerate}
        \item \texttt{main.py} -- Main application entry point
        \item \texttt{database.py} -- Database configuration
        \item \texttt{models.py} -- SQLAlchemy models
        \item \texttt{schemas.py} -- Pydantic validation schemas
        \item \texttt{prompts.py} -- Gemini AI system prompts
        \item \texttt{requirements.txt} -- Python dependencies
    \end{enumerate}

    \item \textbf{frontend/} -- React application
    \begin{enumerate}
        \item \texttt{src/pages/} -- InputPage and Flowchart 
              components
        \item \texttt{src/components/} -- Sidebar and custom 
              nodes
        \item \texttt{src/hooks/} -- \texttt{useFlowchart}, 
              \texttt{useEditableNodes}
        \item \texttt{src/services/} -- API service layer
        \item \texttt{src/utils/} -- Layout algorithms
    \end{enumerate}

    \item \texttt{.env} -- Environment variables (database URL, 
          Gemini API key)
    \item \texttt{package.json} -- Frontend dependencies
    \item \texttt{README.md} -- Project documentation

\end{enumerate}

%---------------------------------------------------------------
\section{Deployment Configuration}
\label{sec:deployment-config}
%---------------------------------------------------------------

The Automated Diagram Digitization platform is deployed using 
modern cloud hosting services to ensure accessibility, 
scalability, and reliable performance. The application follows 
a distributed deployment architecture in which the frontend 
interface and backend services are hosted on separate cloud 
platforms. This separation improves scalability and allows each 
component to be updated independently. \textit{Refer Table~\ref{tab:deployment}: Deployment Architecture Components.}

\subsection{Deployment Architecture}
\label{subsec:deployment-arch}

The system deployment architecture consists of four primary 
components: the frontend client application, the backend API 
service, the PostgreSQL database, and the external AI service 
used for diagram recognition.

\begin{table}[H]
\centering
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{|l|l|}
\hline
\textbf{Component} & \textbf{Deployment Platform} \\
\hline
Frontend (React + Vite) & Vercel \\
\hline
Backend API (FastAPI) & Render \\
\hline
Database & PostgreSQL \\
\hline
AI Service & Google Gemini API \\
\hline
\end{tabular}
\caption{Deployment Architecture Components}
\label{tab:deployment}
\end{table}

\subsection{Frontend Deployment}
\label{subsec:frontend-deploy}

The frontend application was deployed on the Vercel platform. 
Vercel provides optimized hosting for modern JavaScript 
applications and supports automatic builds and deployments 
through Git integration. Whenever changes are pushed to the 
project repository, Vercel automatically builds the React 
application and deploys it through a global content delivery 
network (CDN), ensuring fast loading times and high 
availability. This deployment model enables continuous 
deployment, allowing new updates to reach users quickly while 
maintaining consistent performance across geographic 
regions.

\subsection{Backend Deployment}
\label{subsec:backend-deploy}

The backend API developed using the FastAPI framework is 
deployed on the Render cloud platform. Render hosts the backend 
service and exposes REST API endpoints that handle image 
analysis, text-to-flowchart generation, and database 
operations. The platform automatically manages HTTPS 
configuration, server scaling, and environment variable 
management, ensuring secure communication between the frontend 
and backend services at all times.

\subsection{Environment Configuration}
\label{subsec:env-config}

Sensitive configuration values such as database connection 
strings and API keys are stored as environment variables within 
the deployment platforms. This ensures that confidential 
information is not stored directly in the application source 
code. The backend service retrieves these values at runtime 
using the \texttt{python-dotenv} library to securely connect 
with the database and external AI services, simplifying 
configuration management across development, testing, and 
deployment environments.

%---------------------------------------------------------------
\section{Security Measures}
\label{sec:security}
%---------------------------------------------------------------

Security is a critical aspect of any web application that 
interacts with external services and stores user data. The 
Automated Diagram Digitization platform implements several 
layers of security to protect API credentials, prevent 
malicious inputs, and ensure safe communication between the 
frontend, backend, and database 
components.

\subsection{API Key Protection}
\label{subsec:api-key}

The Google Gemini API key and PostgreSQL connection string are 
stored exclusively as environment variables on the server side 
and are never exposed in client-side code or version control 
repositories. The backend retrieves these values at runtime 
using the \texttt{python-dotenv} library, ensuring that 
sensitive credentials remain confidential across all 
environments. A \texttt{.gitignore} rule prevents the 
\texttt{.env} file from being committed to the 
repository.

\subsection{CORS Configuration}
\label{subsec:cors}

Cross-Origin Resource Sharing (CORS) is configured in the 
FastAPI backend to restrict API access to approved origins 
only. During development, only \texttt{localhost:5173} and 
\texttt{localhost:3000} are permitted. In production, only the 
verified frontend domain is whitelisted, preventing 
unauthorized third-party websites from making requests to the 
backend API.

\subsection{Input Sanitization and Validation}
\label{subsec:input-validation}

All user-provided inputs -- including uploaded image files, 
text prompts, and JSON data -- are validated before processing. 
Image files are checked for allowed MIME types and size limits 
before being passed to the Gemini API. Text prompts are 
length-validated, and JSON imports are parsed with strict error 
handling to reject malformed or malicious 
payloads.

\subsection{SQL Injection Prevention}
\label{subsec:sql-injection}

All database queries are executed through SQLAlchemy's 
parameterized query interface, which separates SQL commands 
from user-supplied data. This eliminates the possibility of 
SQL injection attacks, as user input is never directly 
interpolated into raw query strings.

\subsection{HTTPS Enforcement}
\label{subsec:https}

Both the frontend (deployed on Vercel) and the backend 
(deployed on Render) are served exclusively over HTTPS. All 
data transmitted between the client and server is encrypted 
in transit, preventing interception attacks and ensuring the 
confidentiality of user inputs and AI-generated flowchart 
data.

%---------------------------------------------------------------
\section{Advantages of Technology Stack}
\label{sec:tech-advantages}
%---------------------------------------------------------------

The technology choices made for this project offer several 
significant advantages in terms of performance, developer 
experience, accuracy, cost-effectiveness, and scalability. 
Each subsection below highlights one dimension of this 
benefit.

\subsection{Performance}
\label{subsec:performance}

FastAPI's asynchronous nature handles concurrent requests 
efficiently. React Flow's virtualization supports large 
diagrams with many nodes. JSONB queries in PostgreSQL provide 
fast access to flowchart data. The combination of asynchronous 
processing and efficient database querying reduces latency 
during diagram generation and editing, resulting in smooth 
user interactions even with complex 
diagrams.

\subsection{Developer Experience}
\label{subsec:dev-experience}

FastAPI's automatic OpenAPI documentation simplifies API 
testing. React's component reusability accelerates UI 
development. Clear separation between frontend and backend 
enables parallel development, allowing developers to debug 
and maintain individual components without affecting other 
parts of the system.

\subsection{Accuracy}
\label{subsec:accuracy}

Custom prompt engineering with the Gemini-Flash AI model achieves approximately 92\% accuracy in informal testing across a sample set of complex flowchart images. Accuracy is defined as the successful recognition of all primary flowchart symbols (rectangles, diamonds, ovals, parallelograms), accurate text extraction with character-level precision exceeding 90\%, and correct mapping of directional relationships between shapes. The custom column-snapping layout algorithm further enhances this accuracy by ensuring that recognized diagrams are presented in a professionally aligned and readable format through automated clustering and edge routing optimization.

\subsection{Cost-Effectiveness}
\label{subsec:cost}

Open-source technologies eliminate licensing costs. The Gemini 
API offers competitive pricing with a free tier suitable for 
development and testing. PostgreSQL provides enterprise-grade 
features without vendor lock-in. Cloud hosting platforms such 
as Vercel and Render offer free tiers suitable for development 
and initial deployment, reducing operational expenses while 
still supporting scalable delivery.

\subsection{Scalability}
\label{subsec:scalability}

The stateless FastAPI backend supports horizontal scaling. 
PostgreSQL JSONB allows schema evolution without requiring 
full migrations as new node types or properties are added. 
React's efficient rendering handles complex diagrams. This 
architecture allows the system to accommodate increasing 
numbers of users and diagrams without major structural 
changes.

%---------------------------------------------------------------
%---------------------------------------------------------------
\section{Containerization}
\label{sec:containerization}
%---------------------------------------------------------------

The Automated Diagram Digitization platform is fully containerized using Docker and Docker Compose to ensure environment consistency across development and production stages. This approach eliminates configuration discrepancies and provides a simplified setup process for the multi-service architecture.

\subsection{Docker and Docker Compose}
\label{subsec:docker}

Docker is used to encapsulate the application into portable containers. The \texttt{docker-compose.yml} configuration orchestrates the system’s primary services into a unified environment. This include the frontend container, which uses a Node.js image to compile the Vite application and serve static assets; the backend container, which utilizes a Python image to install dependencies and run the FastAPI server via Uvicorn; and the database container, based on the official PostgreSQL 16 image with persistent volumes for reliable data storage.
Containerization ensures that any developer can spin up the entire stack with a single \texttt{docker-compose up} command, guaranteeing that all services correctly interact through a localized virtual network.

\section{Conclusion}
\label{sec:tech-conclusion}
%---------------------------------------------------------------

The technological framework described in this chapter provides 
the foundational infrastructure required for the Automated 
Diagram Digitization platform to achieve its functional and 
performance objectives. By strategically selecting 
high-performance tools such as FastAPI for backend efficiency, 
React Flow for interactive diagramming, and Google's Gemini AI 
for multimodal recognition, the platform delivers a responsive 
and reliable user experience. These technologies, supported by 
a scalable PostgreSQL data layer and professional development 
tools, collectively enable the seamless transition from 
physical diagrams to an editable digital environment. The 
architecture not only fulfills current requirements but also 
provides the necessary extensibility for future technical 
enhancements discussed in 
Chapter~\ref{ch:future}.