\renewcommand{\bibname}{References}

\begin{thebibliography}{99}
\addcontentsline{toc}{chapter}{References}

% Research Papers and Journals
\bibitem{ocr-review}
R. Vishnu Vardhan Reddy and Dr. G. Rajeswari. (2025). \textit{Automated Document Processing: Combining OCR and Generative AI for Efficient Text Extraction and Summarization}. IJRTI.

\bibitem{diagram-survey}
Alsharif, O., Alsharif, M., \& Alsharif, K. (2025). \textit{Decoding technical diagrams: A survey of AI methods for image content extraction and understanding}. Information, 17 (2). 

\bibitem{arrow-rcnn}
Schäfer, B., Keuper, M., \& Stuckenschmidt, H. (2021). \textit{Arrow R-CNN for handwritten diagram recognition}. International Journal on Document Analysis and Recognition (IJDAR), 24 (1-2), 3-17. doi: 10.1007/s10032-020-00361-1.

\bibitem{flowchart-code}
Shukla, S., Gatti, P., Kumar, Y., Yadav, V., \& Mishra, A. (2025). 
\textit{Towards Making Flowchart Images Machine Interpretable}. arXiv preprint arXiv:2501.17441.

\bibitem{ai-education}
Zawacki-Richter, O., Marin, V. I., Bond, M., \& Gouverneur, F. (2019). \textit{Systematic review of research on artificial intelligence applications in higher education}. International Journal of Educational Technology in Higher Education, 16 (1), 1-27.

\bibitem{rest}
Fielding, R. T. (2000). \textit{Architectural Styles and the Design of Network-based Software Architectures}. UC Irvine Dissertation.

\bibitem{diagram-recognition}
Liu, J., \& Zhang, H. (2023). \textit{Computer Vision-Based Diagram Recognition: A Survey}. Pattern Recognition Letters, 165, 112-124.

\bibitem{flowchart-analysis}
Kim, S., \& Lee, J. (2022). \textit{Automatic Flowchart Recognition from Hand-Drawn Images}. IEEE Access, 10, 45678-45692.

\bibitem{tensorflow}
Abadi, M., et al. (2016). \textit{TensorFlow: A System for Large-Scale Machine Learning}. 12th USENIX Symposium on Operating Systems Design and Implementation, 265-283.

\bibitem{mobilenet}
Sandler, M., Howard, A., Zhu, M., Zhmoginov, A., \& Chen, L. C. (2018). \textit{MobileNetV2: Inverted Residuals and Linear Bottlenecks}. Proceedings of the IEEE Conference on Computer Vision and Pattern Recognition, 4510-4520.

\bibitem{opencv}
Bradski, G. (2000). \textit{The OpenCV Library}. Dr. Dobb's Journal of Software Tools.

\bibitem{szeliski}
Szeliski, R. (2022). \textit{Computer Vision: Algorithms and Applications} (2nd ed.). Springer.

% Software Engineering and Testing
\bibitem{clean-arch}
Martin, R. C. (2017). \textit{Clean Architecture: A Craftsman's Guide to Software Structure and Design}. Prentice Hall.

\bibitem{sommerville}
Sommerville, I. (2016). \textit{Software Engineering} (10th ed.). Pearson.

\bibitem{pressman}
Pressman, R. S., \& Maxim, B. R. (2020). \textit{Software Engineering: A Practitioner's Approach} (9th ed.). McGraw-Hill.

\bibitem{softwaretesting}
Myers, G. J., Sandler, C., \& Badgett, T. (2011). \textit{The Art of Software Testing} (3rd ed.). John Wiley \& Sons.

% Backend Technologies and Documentation
\bibitem{gemini}
Google DeepMind. (2024). \textit{Gemini API - Multimodal Language Model Platform}. Retrieved from https://ai.google.dev

\bibitem{fastapi}
Ramirez, S. (2024). \textit{FastAPI Documentation}. Retrieved from https://fastapi.tiangolo.com

\bibitem{python}
Python Software Foundation. (2024). \textit{Python 3.9 Documentation}. Retrieved from https://docs.python.org/3.9/

\bibitem{uvicorn}
Encode. (2024). \textit{Uvicorn - ASGI Server Documentation}. Retrieved from https://www.uvicorn.org

\bibitem{pydantic}
Colvin, S., et al. (2024). \textit{Pydantic - Data validation using Python type hints}. Retrieved from https://docs.pydantic.dev

\bibitem{google-genai-sdk}
Google. (2024). \textit{Google GenAI Python SDK}. Retrieved from https://github.com/google/generative-ai-python

% Frontend Technologies and Documentation
\bibitem{react}
React Team. (2024). \textit{React 19 Documentation}. Retrieved from https://react.dev

\bibitem{vite}
Vite Team. (2024). \textit{Vite Documentation}. Retrieved from https://vitejs.dev

\bibitem{reactflow}
webkid. (2024). \textit{React Flow Documentation}. Retrieved from https://reactflow.dev

\bibitem{lucide}
Lucide Icons. (2024). \textit{Lucide React Documentation}. Retrieved from https://lucide.dev

\bibitem{react-router}
Remix Software. (2024). \textit{React Router Documentation}. Retrieved from https://reactrouter.com/en/main

\bibitem{react-icons}
React Icons Team. (2024). \textit{React Icons Documentation}. Retrieved from https://react-icons.github.io/react-icons/

\bibitem{html-to-image}
Bubko, Y. (2024). \textit{html-to-image Documentation}. Retrieved from https://github.com/bubkoo/html-to-image

% Database and Storage
\bibitem{postgresql}
PostgreSQL Global Development Group. (2024). \textit{PostgreSQL 16 Documentation}. Retrieved from https://www.postgresql.org/docs/16/

\bibitem{sqlalchemy}
SQLAlchemy Authors. (2024). \textit{SQLAlchemy 2.0 Documentation}. Retrieved from https://www.sqlalchemy.org

\bibitem{jsonb}
PostgreSQL Documentation. (2024). \textit{JSON Types}. Retrieved from https://www.postgresql.org/docs/16/datatype-json.html

% Standards and Tools
\bibitem{cors}
WHATWG. (2024). \textit{Cross-Origin Resource Sharing Specification}. Retrieved from https://fetch.spec.whatwg.org/\#http-cors-protocol

\bibitem{git}
Chacon, S., \& Straub, B. (2014). \textit{Pro Git} (2nd ed.). Apress.


\bibitem{npm}
npm. (2024). \textit{npm Documentation}. Retrieved from https://docs.npmjs.com

% Deployment and Infrastructure
\bibitem{docker}
Docker Inc. (2024). \textit{Docker Documentation}. Retrieved from https://docs.docker.com

% Machine Learning Libraries
\bibitem{pillow}
Clark, A. (2024). \textit{Pillow (PIL Fork) Documentation}. Retrieved from https://python-pillow.org

\end{thebibliography}
